---
title: "ECON 590 Class 1 Activity"
author: "Taylor Mackay"
date: '2023-01-21'
output: html_document
---

## Overview

The goal of this first activity to familiarize ourselves with `RStudio`. We'll learn how to:

-   Install and load packages (collections of `R` commands or functions) needed for data processing

-   Load a sample data set and explore it visually

-   Perform basic data manipulation tasks using `tidyverse`

-   Calculate summary statistics for our data

-   Report our results using both the console and Quarto `.qmd` files

### Setting the Stage

`ECON-590-class-1-activity.qmd` is a `markdown` file - that means you can include regular, formatted text in addition to code and output in the same file.

-   You can run code "in-line" so that results show in the console -- you can either copy and paste the line you want to run into the console below, or select it and press `ctrl + Enter`

    -   You'll see the results from the command printed below on the console

-   You can also click "Render" (or press `CTRL + SHIFT + Enter`) to have an HTML document created with both the written text from the assignment and all your code + output

    -   The resulting file should open in your web browser and will be saved as an `.HTML` file wherever you saved your `markdown` file

### Some `R` Basics

In `R`, we use `<-` as an *assignment operator --* it basically means the same thing as `=` . For instance, entering `x <- 1` creates a new object named `x` that has the numeric value 1. As a shortcut, you can press `ALT + -` instead of having to type out `<-`.

Take a look at the code below to get a sense of how this process works

```{r}

# In R, any line that starts with a # is a comment -- this is how we'll explain what we're doing in a given section of code. Let's start by creating an object named "a" that stores the number 5 

a <- 5

# Now that we've run the code above, we can just type "a" into the command line, and R will return the value 5

a 

# We can also use "a" in an equation like the following 

a + 5 

# You can also create vectors of data in the same way

b.vector <- c(1, 2, 3)

b.vector

# Take a look at the "Environment" pane on the right. Where are "a" and "b.vector" saved?
```

### Writing Readable Code

*Documenting* your code means explaining what you're doing so that someone can read through what you've written and understand what the code does. It is *essential* for producing clear, usable.

-   The ***key difference*** between the programming you might have done in class and what you'll have to do professionally is ***readability*** -- making sure that other people can understand your code

-   We'll talk about how to write clear, readable code throughout this course, but the first step is to include comments in your code (see the code section above for an example)

## Installing and Loading Packages in `R`

"Basic `R`" comes with lots of commands installed already, but one of the benefits of `R` is the wide array of other programs that we can use. Packages in `R` are just collections of commands or functions that we can use to conduct data analysis tasks that aren't included in Base `R`.

Running the code below will install the data processing package `tidyverse` and load it so that you're able to use the commands included with it.

```{r}

# Start by installing the program

# install.packages("tidyverse")

# Load the program so we're able to use the commands included with the package

library(tidyverse)

```

## Load a Sample Data Set

`R` is very flexible in how you can access and load data sets. As we go through the course, we'll cover a range of ways of loading data in `R`. In the example below, we'll load a `.csv`-formatted file that I've saved to GitHub.

-   The function `read.csv()` tells `R` we want to load a `.csv` file

-   The function `url()` tells `R` to go to the URL entered in quotes and look for information to retrieve - this is a handy way of accessing data that's not stored on your local computer.

```{r}

# Load .csv file stored via course GitHub

sample.data <- read.csv(url("https://raw.githubusercontent.com/mackaytc/econ-590-resources/main/data/ACS-4k-obs-sample-data.csv"))

```

### Sample Data Description

Our sample data set `sample.data` contains data from the American Community Survey (ACS). Each of the 4,000 rows in the data set corresponds to a particular person who responsed to the survey -- we'll refer to these rows as ***observations***. Each observation in this data set is then a person. Each column of the data set corresponds to a particular ***variable*** which stores information about each person in the survey. Key variables include:

-   `year`: Tells you the year in which this person or observation took the ACS

-   `statefips`: Tells you the state in which that person lived when they took the ACS

-   `incwage`: Tells you the annual wage and salary information for that person

## Working with Data using `dplyr`

Now we want to demonstrate several data cleaning tasks using `dplyr` functions. We'll start by removing variables from our data set, then demonstrate how to create a ***subset*** of the data (a limited sample of observations).

```{r}

# We want to make the data set a bit smaller and more manageable for today. Let's start by using the select command to keep only the "sample.data", "year", "statefip", and "incwage" variables

sample.data <- select(sample.data, year, statefip, incwage) 

# The head() command will show us a snapshot of what the data set looks like

head(sample.data)

# Suppose we wanted to subset the data so that we only kept observations that lived in New York. To do this, we can use the filter() function to create a new object "sample.data.NY" that stores our subsetted data. 

sample.data.NY <- filter(sample.data, statefip == "new york")

# Take a look at the new data set stored in the "Environment" window. Are all the observations from New York? 

head(sample.data.NY)

# We can also use filter() to select observations by numeric values. This time, we won't create a new object - we'll just output the result using head(). We can use the n = 10 option to show 10 rows of the subsetted data. 

head(filter(sample.data, incwage > 20000), n = 10)

```

## Practice Problems

Using our sample data set, answer the following questions. Before you get started on the activity, take a look at the `dplyr` cheat sheet posted on the course GitHub page.

1)  Using the `mean()` function, report the average wage income for everyone in our sample

```{r}

# Insert your code here

```

2)  Using the `filter()` function, report the average wage income for everyone in our sample who lives in Florida

```{r}

# Insert your code here

```

3)  Report the average wage for everyone who *didn't* live in Florida.

*HINT:* We can use `filter()` again -- try entering `filter(sample.data, statefip %in% "florida")`.

```{r}

# Insert your code here

```

4)  Report the average wage income from observations who who surveyed in 2004 through 2008.

*HINT:* You can use `&&` to add multiple restrictions in your `filter()` command.

```{r}

# Insert your code here

```

5)  Using the `dplyr` cheat sheet on GitHub, calculate the median, minimum and maximum values of the `incwage` variable

```{r}

# Insert your code here

```

6)  Calculate the average wage for each year in the data.

*HINT:* Take a look at the `Group Data` section of the `dplyr` cheat sheet. We want to group the data by year, then calculate average wages using the `summarise()` command.

```{r}

# Insert your code here

```
