---
title: "ECON 590 Week Two Handout"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Basic Data Structures in `R`

Last class, we said that `<-` is the assignment operator. We created an object 
named `a` that stored the value `4` using the following command. 
```{r Assignment, echo = T}

a <- 4

```

In `R`, we can create several different kinds of objects or data types using 
the assignment operator: 

 * __Numeric__: stores a single value (like `a` stores the number `4` above)
 * __String__: stores a string of text (like "hello world")
 * __Logical__: stores binary values True or False
 * __Factor__: stores categories (like "Employed," "Employed Part-Time," 
 and "Unemployed)

```{r Data Types, echo = TRUE}

# Below are examples of various data types and functions to identify them. 
# First, let's check if a is numeric. `is` functions will always return 
# true or false.

a <- 4 

is.numeric(a)

# Now we can create a string object and check it using is.character()

a.string <- "123 Main St"

is.character(a.string)

# When we create a logical object, note that we don't put TRUE in quotes
# (otherwise it would be a string object)

a.logical <- TRUE

is.logical(a.logical)

# We'll take a longer look at factor variables in a future class

```
# Vectors 

We also showed last class how we can create vectors storing values, like 
the following: 
```{r Vector Assignment, echo = T}

b.vector <- c(1, 2, 3)

```
In the code above, the `c()` function _combines_ or _concatenates_ the 
arguments in parentheses. Vectors are the basic data structure in `R`, so we
want to make sure we understand how they work: 
```{r Vector, echo = T}

# Basic operations in R get applied to each element of a vector

b.vector + 2 

# We can access individual elements in a vector using []'s

b.vector[2]

b.vector[2:3]

```
# Data Frames in `R`

When we work with data in `R`, we'll generally be using `data frames`. 
Technically, data frames are just a collection or list of vectors. We can 
create a sample data set below. 

```{r Data Frames, echo = TRUE}

# Define vectors storing various object types from above

person.ID <- c(12, 24, 54, 65)
address   <- c("123 Main St.", "274 Long St.", "789 Right St.", "467 Left St.")
employed  <- c(TRUE, TRUE, FALSE, TRUE)
wage.inc  <- c(12500, 15750, 0, 14100)

# Combine each of the individual vectors into a data frame

data <- data.frame(person.ID, address, employed, wage.inc)

```

You might know what each of the object types defined above are already, but
you can check the "Environment" pane to confirm. Find our `data` object, and 
click the triangle button to see more information. We can access individual
variables (or vectors) in the data frame using the `$` operator: 

```{r Accessing Variables in Data Frames, echo = TRUE}

# Code below prints the entire person.ID vector then just particular rows

data$person.ID

# We can also perform operations on the individual variables

mean(data$employed) # Note that T = 1, F = 0 here
mean(data$wage.inc)

```
# Accessing Data in Data Frames

There's two general ways of retrieving data from data frames. Generally, we'll
want to use `tidyverse` commands where possible for larger data sets (more on
that in the class activity for today). But its important to know the `base R` 
approach as well. In the section above, we showed we can use `$` to pick out 
specific variables or vectors. 

In this section, we'll show several ways of accessing data within data frames.

```{r Indexing, echo = TRUE}

# We can use brackets to access data within a particular variable. The command
# below returns the first address in our data set

data$address[1]

# We could also select multiple observations sequentially using the following

data$address[2:4]

# Finally, you can provide a list of row indexes like so 

data$address[c(1,3,4)]

# One useful function to learn about the structure of a data frame is str()

str(data)

```
