---
title: "ECON 590: Class 2 Activity Solutions"
output: html_notebook
---

# Class 2 Activity Overview

The goal today is to learn about the basic `tidyverse` functions for data 
cleaning and manipulation. Some of the most useful functions are listed below:

-   `select()` - select specific columns or variables from your data
-   `filter()` - select specific rows or observations from your data
-   `arrange()` - reorders or sorts the rows of your data
-   `summarize()` - creates summary statistics (often using `group_by()`)
-   `mutate()` - creates new variables that are functions of existing variables

To do this, we'll use the same ACS data set from last class. You can load this
data set by running the code section below: 

```{r Load Data, echo = TRUE }
# Load the tidyverse package

library(tidyverse)

# Load .csv file stored via course GitHub

data.url <- url("https://raw.githubusercontent.com/mackaytc/econ-590-resources
                 /main/data/ACS-4k-obs-sample-data.csv")

sample.data <- read.csv(data.url)

# Just like last class, we can use select() to keep just the variables we want
# to work with below

sample.data <- select(sample.data, year, state, incwage, foodstmp) 

```

In this activity, we'll look at two variables: 

 * `incwage`: total annual wage income (remember that we want to set any 
      observations of `incwage = 999999` to missing)
 * `foodstmp`: binary indicator for receiving food stamps / SNAP (coded as 
      a string variable with `"yes"` and `"no"` responses)

# Questions

For each question below, provide your response in the `R` code block. Make
sure to include comments so its clear what your code is intended to do!

### Part 1: Use `mutate()` to update the `incwage` variable to account for missing observations (see above)

After you've created missing values of `incwage`, use `mean(is.na())` to check
what fraction of observations have a missing value for `incwage`.

```{r Data Cleaning incwage, echo = TRUE}
# Use the code from last week's activity (posted to Canvas) to enter missing
# values for the incwage variable using the mutate() & replace() functions

sample.data <- mutate(sample.data, 
                      incwage = replace(incwage, incwage == 999999, NA))

mean(is.na(sample.data$incwage))

```
### Part 2: Create a new variable that's coded as a logical variable equal to `TRUE` when the survey respondent reports receiving food stamps / SNAP via the `foodstmp` variable

Use the `mean()` function again to calculate what fraction of observations
received food stamps / SNAP benefits.

```{r Data Cleaning foodstmp, echo = TRUE}
# You can use mutate() here again -- try using / Googling ifelse() as well

sample.data <- mutate(sample.data, 
                      foodstmp.logical = ifelse(foodstmp == "yes", 1, 0))

mean((sample.data$foodstmp.logical))

```

### Part 3: Create a new variable with the log-value of incwage (via the `log()` function)

You might've noticed that there's lots of 0's in the data! Remember that 
`log(0)` isn't defined. There's several things you could do here depending on 
the context, but for now, you can just add `1` to `incwage` then calculate 
the logged value.

```{r Data Cleaning Log Wages, echo = TRUE}
# INSERT your code below

sample.data <- mutate(sample.data,
                      log.incwage = log(incwage + 1))

```

### Part 4: Calculate average income and food stamp recipiency rates by year

HINT: We did something similar using `group_by()` and `summarize()` as part
of last week's activity. Make sure you're using the new, logical version of 
the food stamps variable. 

```{r Data Cleaning Summary Stats by Year, echo = TRUE}
# INSERT your code below

summary.stats <- group_by(sample.data, year) %>% 
  summarize(avg.annual.incwage  = mean(incwage, na.rm = TRUE),
            avg.annual.foodstmp = mean(foodstmp.logical))

summary.stats
```

### Part 5: Create a new data frame that stores average food stamp recipiency rates for each state _for the last year in the sample only_ and use then use the `arrange()` function to sort this new data frame from highest to lowest rates

NOTE: We don't have all that many observations for each state + year, so 
don't worry if your data looks a bit wonky at the end here (your highest 
recipiency rate state should have a rate of 100 percent). 

```{r Data Cleaning SNAP Recipiency, echo = TRUE}
# INSERT your code below

avg.state.foodstmp <- filter(sample.data, year == 2019) %>% 
  group_by(state) %>% 
  summarize(avg.state.foodstmp.rates = mean(foodstmp.logical)) %>% 
  arrange(desc(avg.state.foodstmp.rates))

head(avg.state.foodstmp)

```
